\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{xcolor}
\usepackage{tikz}

\title{Evolution of communication in populations of honeybee-like agents}
\author{Grzegorz Chrupa{\l}a}
\date{\today}

\begin{document}

\maketitle

\section{Introduction}

Honeybee waggle dances provide spatial information about resources away from
the nest. One ecological hypothesis is that such communication is favored when
food is difficult to discover independently but remains useful after a worker
has found it. Empirical and theoretical studies suggest that the value of dance
information and recruitment depends on the spatiotemporal distribution of food
resources, including patch density, patch size, reward variability, and habitat
\cite{sherman_visscher_2002,dornhaus_chittka_2004,dornhaus_etal_2006,beekman_lew_2008,donaldson_matasci_dornhaus_2012,schurch_gruter_2014,price_gruter_2015}.
The model is deliberately minimal: it asks which evolutionary pressures are
needed before adding richer social, developmental, or landscape detail. We start
from direct directional communication on an exposed horizontal comb, where dance
angle can point in the food direction, and then ask what additional conditions
make a gravity-referenced vertical-comb code reachable.

The simulations evaluate four evolutionary hypotheses.
\begin{enumerate}
\item \textbf{Resource-distribution hypothesis.} Costly directional
communication should be favored when independent discovery is difficult enough
that recruitment is useful, but not so difficult that few dances are seeded.
It should be weakly favored when food is very easy to find without dances.
\item \textbf{Architectural-precondition hypothesis.} Direct pointing should
work well on a horizontal comb, whereas gravity-referenced communication should
become useful only when the comb geometry supplies a reliable gravity cue.
A non-communication benefit of vertical combs may therefore be needed to make
the gravity channel selectively available.
\item \textbf{Sender--receiver coordination hypothesis.} Sender and receiver
transposition must evolve together. Positive coupling between sender and
receiver mutations should reduce the time spent in mismatched codes.
\item \textbf{Transition-risk hypothesis.} Selection for vertical combs can
outpace adaptation of the communication code. Successful transitions should
therefore require overlap among vertical geometry, directional precision,
receiver attention, and matched sender--receiver transposition; otherwise
foraging may collapse before the new code becomes useful.
\end{enumerate}

\section{Model}

\subsection{Conceptual model}

Colonies are the reproducing entities, while workers are behavioral agents
sampled from colony-level trait means. A colony has seven heritable mean traits:
directional precision investment, receiver attention, sender transposition,
receiver transposition, search limit, comb tilt, and comb orientation.
Directional precision investment is a costly investment in producing
directionally informative dances. Receiver attention controls whether workers
use available dances. Sender and receiver transposition interpolate between two
communication modes: direct pointing in the comb plane and sun/gravity-referenced
encoding. Comb tilt and orientation define the dance plane in the
world, and search limit determines how far a worker searches along its chosen
direction.

Each foraging episode samples food sites with direction, distance, angular
width, value, and capacity. Workers act sequentially. If dances are available,
a worker may follow one; otherwise it searches independently in a random
direction. Independent searchers that find food add a noisy dance for that site.
There are therefore no fixed scouts in the model. Background discovery and
recruitment both emerge from the food distribution, receiver attention, search
limits, and dance precision. This follows ecological recruitment models in
which communication is valuable only when resources are discoverable and worth
recruiting to \cite{dornhaus_etal_2006,beekman_lew_2008}.

Comb geometry determines which directional cue is available. Direct pointing
projects the horizontal food direction onto the comb plane, so its strength
depends on the food direction, comb tilt, and comb orientation. Gravity-
referenced mapping uses the projection of gravity onto the comb plane together
with the episode's sun azimuth. Gravity provides no in-plane reference on a
horizontal comb and becomes a stronger cue as the comb becomes vertical.
Comb orientation can be treated as circular, or as axial when orientations
$\phi$ and $\phi + 180^\circ$ represent the same comb plane.

\subsection{Mathematical implementation}

Let a colony's mean strategy be
\[
\theta = (b, a, s, r, \ell, t, \phi),
\]
where $b \in [0,1]$ is directional precision investment, $a \in [0,1]$ is
receiver attention, $s,r \in [0,1]$ are sender and receiver transposition,
$\ell \geq 0$ is search limit, $t \in [0,1]$ is comb tilt, and $\phi$ is comb
orientation. A worker $i$ receives stable within-colony deviations from the
colony means:
\[
\theta_i = \operatorname{clip}(\theta + \epsilon_i).
\]
These deviations represent worker heterogeneity but are not themselves
heritable.

For a food direction $\alpha$ and sun azimuth $\psi$, let $D(\alpha;t,\phi)$ be
the direct projected dance angle and $G(\alpha,\psi;t,\phi)$ be the
sun/gravity-referenced dance angle. Let $q_D$ and $q_G$ be their corresponding
cue strengths in the comb plane. A sender with transposition $s_i$ encodes the
dance mean angle as a weighted circular mean,
\[
\mu_i =
\operatorname{cmean}\left(
  (D, (1-s_i)q_D),\,
  (G, s_i q_G)
\right).
\]
The produced dance signal is
\[
x_i =
\operatorname{VM}(\mu_i, \kappa_{\max} b_i) + \eta_d
\pmod{2\pi},
\]
where $\operatorname{VM}$ is the von Mises distribution, $\kappa_{\max}=14$ in
the reported simulations, and $\eta_d$ is transient dance-production noise.
A receiver with transposition $r_j$ applies the analogous weighted decoding
rule and adds interpretation noise $\eta_r$ to obtain a search direction
$\hat{\alpha}_j$.

A worker succeeds at food site $k$ only if
\[
\Delta(\hat{\alpha}_j,\alpha_k) \leq w_k
\quad\text{and}\quad
d_k \leq \ell_j,
\]
where $\Delta$ is circular angular distance, $w_k$ is site angular width, and
$d_k$ is site distance. If multiple available sites satisfy this condition, the
nearest site is used.

Let the net foraging payoff for an episode be
\[
F_e =
\sum_{j \in \mathcal{F}_e} v_{k(j)}
- c_T\left(\sum_{j \in \mathcal{F}_e} d_{k(j)}
  + \sum_{j \in \mathcal{U}_e} \ell_j\right)
- \sum_{i \in \mathcal{D}_e} (c_0 + c_b b_i)
- c_a A_e .
\]
Here $\mathcal{F}_e$ is the set of successful workers, $\mathcal{U}_e$ the
unsuccessful workers, $\mathcal{D}_e$ the independent finders that produce
dances, and $A_e$ the number of dance-following attempts. The terms are food
value, travel/search cost, per-dance communication cost, and receiver-attention
cost. A vertical comb can then have an optional non-communication advantage
that scales, rather than adds to, net performance:
\[
P_e = F_e \left(1 + \alpha f(t)\right).
\]
The default modifier is $f(t)=t$, so $\alpha=0.05$ means that a fully vertical
comb has a 5\% advantage over a horizontal comb with the same foraging
performance. The long-transition robustness check also uses
$f(t)=\mathbf{1}[t \geq 0.8]$. Because the advantage is multiplicative,
verticality cannot rescue a colony whose foraging payoff has collapsed. Mean
colony payoff is averaged over episodes and floored at 0.001 so that low-payoff
colonies remain selectable but rare.

Daughter colonies are sampled in proportion to payoff. Only colony means are
inherited. Mutations are Gaussian and clipped to the allowed ranges; sender and
receiver transposition mutations can have correlation $\rho$, which is the
coupling parameter used in the vertical-transition probes.

\subsection{Simulation parameters and reproducibility}

Unless otherwise stated, reported runs use 60 colonies, 80 workers per colony,
50 foraging episodes per colony per generation, 12 foraging attempts per
episode, ten random seeds, within-colony standard deviation 0.08 for
unit-interval traits, mutation standard deviation 0.07, dance-production noise
standard deviation 0.18, interpretation-noise standard deviation 0.12, maximum
search distance 8.0, food value 1.0, food-site capacity 6, travel cost 0.02
per distance unit, precision-dependent dance-cost coefficient $c_b=0.02$,
baseline dance cost $c_0=0$, and attention cost $c_a=0.01$. Search-limit noise
and mutation use the same scales multiplied by the maximum search distance.
Initial directional precision is sampled from $[0,0.15]$, initial receiver
attention from $[0,0.25]$, initial transposition traits are 0, and initial
search limit is sampled from 15--45\% of the maximum search distance.

The tables and figures in this report are generated from tracked result files
rather than hand-entered into the document. Running
\[
\texttt{python -u experiments/run\_report\_artifacts.py artifacts}
\]
rebuilds the LaTeX fragments from the CSV files in \texttt{results/}; running
the same script with \texttt{all} first regenerates the result CSVs. The
provenance manifest is \texttt{report/artifact\_sources.csv}.

\section{Results}

\subsection{Horizontal comb}

We first compared three simple food distributions over ten random seeds
(101--110). The ``hard'' condition had one narrow food site per episode, the
baseline condition had two moderately wide food sites, and the ``easy''
condition had many broad food sites. Table~\ref{tab:food-distribution} reports
the generation at which mean directional precision investment first reached
0.30, plus final-generation trait, success, and payoff values.

Each run used 60 colonies, 80 workers per colony, 30 generations, 50 foraging
episodes per colony per generation, 12 foraging attempts per episode, and
horizontal-comb tilt $t=0$. Food-site distances were sampled uniformly from
1.0 to 8.0, maximum search distance was 8.0, travel cost was 0.02 per distance
unit, baseline dance cost was 0, the precision-dependent dance-cost coefficient
was 0.02, and food-site value was 1.0. All three conditions used food-site
capacity 6. The conditions varied only the number and angular width of food
sites: hard used one site of width 0.08, baseline used two sites of width 0.20,
and easy used eight sites of width 0.50. The hard condition therefore makes
independent discovery unlikely, while the easy condition makes independent
discovery likely even without using the dance.

\input{tables/food_distribution.tex}

The result supports the resource-distribution hypothesis. In the hard
environment, food is so rare and narrow that independent discoveries seldom
seed dances; final success and payoff remain close to zero, and communication
crosses the directional-precision threshold in only 4 of 10 seeds. In the
baseline environment, independent discovery is sufficient to seed recruitment
cascades, and directional precision, receiver attention, and search limit all
rise strongly. In the easy environment, success and payoff are high, but
directional precision and receiver attention remain much lower than in the
baseline case because food is often found without precise communication.

\subsection{Vertical comb}

We also compared colonies initialized with horizontal, tilted, and vertical
combs while holding the baseline food distribution fixed. Initial comb tilt was
set to $t=0$, $t=0.5$, or $t=1.0$, with ten seeds per condition. Comb tilt was
still heritable and mutable in these runs. Table~\ref{tab:comb-tilt} summarizes
the result after 30 generations.

\input{tables/comb_tilt.tex}

With mutable tilt, all three initial-tilt conditions reach the
directional-precision threshold in all ten seeds under the current baseline
settings. Tilted and vertical starts retain substantial foraging success over
30 generations, but they also evolve higher sender and receiver transposition
than horizontal starts. The initially vertical condition still has lower payoff
than the horizontal condition, and its final mean tilt is reduced in the
underlying result file, indicating that the short-run comparison does not by
itself establish a stable vertical-comb transition.

\subsection{Tilt-geometry calibration}

We next ran a calibration grid over initial comb tilt and the proportional
vertical-comb advantage $\alpha$ using the default modifier $f(t)=t$. These
runs used the baseline food distribution, the daytime sun arc from $0^\circ$
to $180^\circ$ centered at $90^\circ$, and ten seeds (101--110). Success is
the fraction of foraging attempts that found food. Payoff is the final mean
colony payoff after food rewards, search/travel costs, dance costs, attention
costs, and the multiplicative vertical-comb modifier. Table~
\ref{tab:tilt-benefit-calibration} shows that proportional advantages of
15--25\% do not by themselves make horizontal or tilted starts retain
verticality within 30 generations, but vertical starts can retain more tilt.

\input{tables/tilt_benefit_calibration.tex}

The calibration suggests that the transition is strongly path-dependent.
Starting from a horizontal or tilted comb, the tested advantages leave final
mean tilt below 0.25 while preserving high foraging success. Starting from a
vertical comb, final tilt is higher, especially at $\alpha=0.15$ and
$\alpha=0.25$, but success and payoff are lower than in flatter conditions.
The proportional rule therefore removes the artificial life-support effect of
an additive bonus: verticality is retained only when enough foraging
performance remains to be scaled. The relevant open question is whether
longer runs can turn supplied verticality into a coordinated gravity-referenced
code before selection drives colonies back toward flatter combs.

\subsection{Tilt orientation}

We also examined the evolved orientation of the comb tilt. The current sun arc
is centered at $90^\circ$, so Table~\ref{tab:orientation-calibration} reports
the weighted mean orientation in degrees and its offset from this sun-arc
center. The within-alignment column measures how strongly colonies within a run
converge on an orientation. The across-alignment column measures whether
different seeds converge on the same absolute orientation, weighting each seed
by its within-run orientation alignment.

\input{tables/orientation_calibration.tex}

Orientation sometimes aligns within a run, but it does not show a clean,
robust relationship to the daytime sun arc across seeds under the circular
orientation representation. Some conditions point near the sun-arc center, some
near the opposite direction, and several have weak across-seed alignment. This
indicates weak or path-dependent selection on orientation in these probes and
motivates the axial representation used in the longer transition runs, where
$\phi$ and $\phi + 180^\circ$ are treated as the same comb plane.

\subsection{Constrained vertical coupling}

The tilt-calibration results left open whether gravity-referenced
communication is difficult because the sender--receiver code itself is hard to
reach, or because colonies must evolve the vertical geometry and the
sender--receiver code at the same time. To separate these possibilities, we ran
a constrained probe with the geometric precondition supplied. Initial comb tilt
was fixed near vertical at $t=0.95$, comb-tilt mutation was set to 0, comb
orientation was treated axially, the vertical-comb advantage was set to 0, and
the baseline food distribution was otherwise retained. We then varied only the
sender--receiver transposition mutation correlation and ran each condition for
120 generations over five seeds (101--105). A run was counted as reaching
gravity-referenced communication when both mean sender and receiver
transposition reached 0.50.

\input{tables/vertical_coupling_probe.tex}

Table~\ref{tab:vertical-coupling-probe} separates reachability from transition
speed. With near-vertical geometry supplied, every correlation condition
reaches gravity-referenced communication in all five seeds, so the gravity
channel is not intrinsically unreachable. The main difference is how quickly
the matched sender--receiver code appears. Independent or weakly coupled
mutations ($\rho=0$ or $0.3$) reach the threshold only after about 43--46
generations on average. Moderate to strong coupling ($\rho=0.6$--$1.0$)
reaches it by about generation 21--23. The final sender--receiver gap also
falls as coupling increases: from 0.033 without coupling to 0.010 at
$\rho=0.6$ and 0 at $\rho=1.0$. Payoff peaks at $\rho=0.6$, not at perfect
coupling, suggesting that coupling helps align the code but complete lockstep
is not necessarily the highest-payoff mutation structure in this finite-seed
probe. The result supports the sender--receiver coordination hypothesis and
implies that the harder problem is the joint transition in which comb
architecture, orientation representation, and the sender--receiver code must
all become useful together.

\subsection{Long transition runs}

We therefore ran longer transition experiments using the axial orientation
representation. These runs used the baseline food distribution, 120
generations, ten seeds (101--110), sender--receiver transposition mutation
correlation 0.6, three initial comb tilts ($0$, $0.5$, and $1.0$), and
proportional vertical-comb advantages $\alpha \in \{0.05,0.10,0.15,0.20,0.25\}$.
The default runs used $f(t)=t$, and a robustness check used
$f(t)=\mathbf{1}[t \geq 0.8]$. Unlike the constrained near-vertical probe,
comb tilt was free to mutate. A seed was counted as reaching the gravity
channel when both mean sender and receiver transposition reached 0.50. It was
counted as retaining verticality when final mean comb tilt was at least 0.80.
It was counted as collapsed when mean foraging success fell to 0.02 or below
at any generation, and as recovered when a collapsed seed ended with final
success at least 0.10.

\input{figures/long_transition_heatmap.tex}

The proportional rule changes the interpretation of the long runs
(Figure~\ref{fig:long-transition-heatmap}). From horizontal starts, no tested
advantage reaches the gravity channel or retains verticality under either
modifier, and final success remains high. From tilted starts, no tested
advantage reaches the gravity channel or retains verticality either; a few
seeds transiently collapse but recover by the final generation. Thus
proportional verticality pressure up to 25\% is not sufficient to bootstrap the
architecture from horizontal or intermediate tilt under the current mutation
and ecological settings.

From vertical starts, by contrast, verticality and gravity-referenced
communication sometimes coevolve without lasting foraging failure. With
$f(t)=t$, the number of seeds retaining verticality and reaching the gravity
channel is 2/10 at $\alpha=0.05$, 2/10 at $\alpha=0.10$, 4/10 at
$\alpha=0.15$, 0/10 at $\alpha=0.20$, and 3/10 at $\alpha=0.25$. With the
threshold modifier, the corresponding counts are 2/10 for
$\alpha=0.05$--$0.20$ and 3/10 at $\alpha=0.25$. Collapse is limited to one
seed in each vertical-start condition and those seeds recover by generation
120. The transition is therefore possible when vertical architecture is already
supplied, but it remains stochastic and non-monotonic rather than robust.

\subsection{Food-distribution transition probe}

The long transition runs showed that the baseline food distribution does not
bootstrap verticality from horizontal or tilted starts. We next asked whether
more forgiving food distributions could create a viable transition corridor
from horizontal starts without changing the inheritance or mutation rules. All
runs in this probe started at $t=0$, used axial orientation, retained the
sender--receiver mutation correlation $\rho=0.6$, used the proportional
vertical-comb modifier $f(t)=t$ with $\alpha=0.25$, and ran for 120
generations over five seeds (101--105). We varied only the food distribution:
angular width, site count, capacity, and value.

\input{tables/food_transition.tex}

The result is mixed but informative. Broader, richer, or higher-capacity
patches substantially improve foraging success relative to the baseline, but
they do not by themselves make the gravity code or stable vertical combs evolve
from horizontal starts. For example, broad high-capacity patches raise final
success to 0.391, but no seed reaches the gravity threshold or retains
verticality, and mean final tilt remains 0.115.

The only successful horizontal-start transition in this probe occurs in the
moderate-dense condition with four sites, angular width 0.30, capacity 10, and
unit value. One of five seeds reaches the gravity-channel threshold at
generation 75, reaches verticality at generation 108, and ends with comb tilt
0.911 and sender and receiver transposition 0.863 and 0.862. The same condition
also has the highest final success (0.470). The other four seeds in that
condition, however, stay far from vertical and do not reach the gravity
threshold. Thus this ecology can open a transition path from horizontal starts,
but the path is still rare rather than reliably selected under the current
settings.

\subsection{Local food-transition grid}

To resolve the neighborhood around the seed-101 transition, we ran a local
grid from horizontal starts while keeping the model rules fixed. The grid used
seed 101, axial orientation, $\rho=0.6$, food value 1.0, $f(t)=t$, 120
generations, and four worker processes. It varied the food-site count
$n \in \{3,4,5\}$, angular width $w \in \{0.26,0.30,0.34\}$, capacity
$K \in \{8,10,12\}$, and vertical-comb advantage
$\alpha \in \{0.20,0.25,0.30\}$, for 81 conditions. Table~
\ref{tab:food-transition-local-grid} lists the only conditions that ended in
a stable vertical gravity-code state.

\input{tables/food_transition_local_grid.tex}

The local grid confirms that the earlier transition is reproducible for the
same seed and parameter setting, but it also shows that the transition is not a
smooth monotone response to easier foraging or stronger verticality pressure.
Only 4 of 81 seed-101 conditions are stable. The original pocket
($n=4,w=0.30,K=10,\alpha=0.25$) remains stable, while its immediate
$\alpha=0.20$ and $\alpha=0.30$ neighbors do not. Three additional stable
conditions occur with five food sites and $\alpha=0.30$. In contrast, many
high-success conditions at width 0.34 remain flat and do not reach the gravity
threshold. The transition therefore appears to require a fairly specific
balance: enough discovery and recruitment opportunity to keep foraging viable,
but not so much ecological tolerance that vertical architecture and
gravity-referenced transposition are weakly selected.

We then reran the four stable pockets over seeds 96--106 for 160 generations.
This seed panel is a reproducibility check rather than a statistical
neighborhood in the random-number sequence. Table~
\ref{tab:food-transition-local-robustness} shows that the pockets remain rare.

\input{tables/food_transition_local_robustness.tex}

All four seed-101 stable states remain stable when extended to generation 160,
so they are not transient generation-120 crossings. Across the 44 robustness
runs, however, only five stable outcomes occur. The strongest pocket is
$n=5,w=0.26,K=8,\alpha=0.30$, which succeeds in seeds 100 and 101. The other
three pockets succeed only in seed 101. None of the robustness runs collapses,
and final success remains high even in non-transition runs, so the limiting
factor is not colony viability alone. Most failures keep foraging performance
while either tilt stays low, transposition stays low, or both. This supports
the view that the ecology can make the transition possible, but the
architecture-and-code lock-in remains stochastic under the current mutation
and selection regime.

\subsection{Adaptive food-transition search}

Manual grids become inefficient once several ecological parameters plausibly
interact. We therefore added an Optuna-based adaptive search over the same
scientific quantities, without introducing new biological mechanisms. The
search starts from the long-transition configuration and keeps horizontal
initial combs, axial orientation, $\rho=0.6$, food value 1.0, $f(t)=t$, and
120 generations. It varies food-site count (3--6), angular width
(0.22--0.36), capacity (6--14), vertical-comb advantage (0.15--0.35),
maximum food distance (6.0--10.0), and travel cost (0.01--0.04). For each
seed, let $t_f$ be final comb tilt and let $s_f,r_f$ be final sender and
receiver transposition. A trial is stable when
$t_f \geq 0.80$, $s_f \geq 0.50$, and $r_f \geq 0.50$. Its continuous
near-miss score is
\[
p = \min\left(1, \frac{t_f}{0.80}, \frac{s_f}{0.50}, \frac{r_f}{0.50}\right).
\]
The optimization objective is stable seed count plus mean $p$, minus the
number of seeds whose success falls to 0.02 or below. Thus an additional stable
vertical gravity-code seed always dominates a merely better near miss, while
the continuous term still lets the optimizer rank incomplete transitions.

As a pilot, we ran eight seed-101 trials with four workers. One trial reached
the target state: six food sites, width 0.35, capacity 13,
$\alpha=0.33$, maximum food distance 8.0, and travel cost 0.01 ended with
$t_f=0.914$, $\min(s_f,r_f)=0.855$, and final success 0.608. The other seven
trials did not reach stability; their progress scores ranged from 0.129 to
0.319. This first adaptive pass therefore finds another viable seed-101
ecology, somewhat outside the earlier three-to-five-site local grid, but it
should be treated as a search result rather than a robustness result until the
best regions are rerun over multiple seeds.

We then continued the same Optuna study for 24 total trials over seeds
100--102, for 72 full seed simulations. No trial was stable in all three
seeds, but the adaptive search did move beyond the single-seed result. The
best trial was stable in two of three seeds, with six food sites, width 0.32,
capacity 9, $\alpha=0.34$, maximum food distance 6.0, and travel cost 0.03.
It had objective 2.761, mean progress 0.761, mean final success 0.589, mean
final comb tilt 0.669, and mean final minimum sender--receiver transposition
0.596. Three additional trials were stable in one of three seeds: the seed-101
pilot setting; a nearby six-site setting with width 0.32, capacity 9,
$\alpha=0.35$, maximum distance 6.0, and travel cost 0.03; and a lower-capacity
six-site setting with width 0.28, capacity 7, $\alpha=0.25$, maximum distance
6.0, and travel cost 0.025. None of the 24 trials collapsed. This suggests a
more specific corridor than ``easier food'' alone: relatively many moderate
patches, short food distances, and non-negligible travel costs can make
vertical gravity-code transitions repeat across more than one seed, but the
transition is still not reliable across the small seed panel.

\section{Conclusion}

The resource-distribution experiments support the idea that costly directional
communication is favored most strongly in an intermediate ecological regime:
food must be hard enough to make recruitment valuable, but discoverable enough
for independent finders to seed dances. The geometry experiments support the
architectural-precondition hypothesis in a more conditional way. A
proportional vertical-comb advantage can sometimes preserve supplied
verticality, and supplied near-vertical geometry allows gravity-referenced
sender--receiver transposition to evolve, but the full architecture-and-code
transition is not yet robust.

The constrained coupling probe supports the sender--receiver coordination
hypothesis: with near-vertical geometry supplied, gravity-referenced
communication evolves in every condition, and correlated sender--receiver
mutations shorten the transition and reduce final mismatch. The long transition
runs support the transition-risk hypothesis under a stricter viability-scaled
fitness rule. Proportional vertical-comb advantages up to 25\% do not bootstrap
verticality from horizontal or tilted starts under the baseline ecology, and
vertical starts produce gravity-referenced vertical outcomes only in a minority
of seeds. The food-distribution probes suggest that ecology can matter: a
moderate-dense environment produced one stable horizontal-to-vertical
gravity-code transition in five seeds, and a local grid around that result
found four seed-101 parameter pockets that sustain the transition. But simply
making patches broader, richer, or higher-capacity mostly increases foraging
success without making the architecture-and-code transition robust. The
robustness panel reinforces this point: the best local pocket succeeds in only
2 of 11 seeds, and the other pockets succeed only in seed 101. The next
modeling priority is therefore to use the adaptive search to map the
ecological corridor more carefully and identify what converts rare viable
paths into reliable selection: wider seed follow-up around the two-of-three
Optuna pocket, local refinement around six-site short-distance conditions,
altered mutation/coupling structure, or additional mechanisms that keep
partially vertical combs available while gravity-referenced communication
becomes useful.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
